---
title: "Logistic Regression Analysis Report"
author: "Grace Hardy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

# Load packages 
library(tidyverse)
library(broom)
library(ggplot2)
library(lmtest)
library(car)
library(caret)
library(MASS)
library(glmnet)
library(remotes)
library(rlang)
library(here)
library(caTools)
library(data.table)

```

```{r fileload, include=FALSE}
# Create file path, the here function enables files to be located relative to the project root
diabetes_path <- here("Data", "diabetes.csv")

#tryCatch statement to safely try loading diabetes data
if (file.exists(diabetes_path)) {
  diabetes_df <- tryCatch(
    read.csv(diabetes_path, stringsAsFactors = FALSE),
    error = function(e) {
      stop("Error reading CSV file: ", e$message)
    }
  )
  cat("Diabetes file loaded successfully.\n")
} else {
  cat("Diabetes file not found in current working directory.\n")
}
```
# 1. Exploratory Analysis

Print the first few rows of the diabetes data:
```{r head, echo=FALSE}
# Use head function to display first 6 rows 
head(diabetes_df)
```

Print the last few rows of the diabetes data:
```{r tail, echo=FALSE}
# Use tail function to display last 6 rows 
tail(diabetes_df)
```
No obvious data or formatting errors and the column headings look representative of their values. 

Summary statistics of the diabetes data:
```{r summary, echo=FALSE}
# Use summary function to display summary statistics of data frame 
 summary(diabetes_df)
```

Display the type of each variable:
```{r type, echo =FALSE}
str(diabetes_df)
```

All variables are numerical so do not need converting to a factor except for the Outcome variable which should be treated as a factor for logistic regression. 

```{r Outcome Variable Conversion, echo=FALSE}
#Define Outcome variable
outcome <- diabetes_df$Outcome

#Convert outcome variable to factor
diabetes_df$Outcome <- factor(diabetes_df$Outcome, levels = c(0, 1), labels = c("No", "Yes"))

#Define Outcome variable
outcome <- diabetes_df$Outcome
```

For pregnancies, maximum value of 17 seems quite large but mean is quite close to the median so no indication of right-skewness and this could be a plausible real-world value.

Zero values for Glucose, BloodPressure, SkinThickness, Insulin and BMI could suggest some missing data collection. This is shown below. 

```{r missing values, echo=FALSE}

#Find all rows that have a value of 0
colSums(diabetes_df == 0)

```

The table above shows that the highest number of zero values occur in the SkinThickness and Insulin variables, further implying that there is some missing data for these values, which may impact the validity of further model applications. Pregnancy is also showing as having zero values, but this is a real-world expected value.

For Insulin, the mean is significantly larger than the median which could indicate right skewness or presence of outlier values, the large max value in comparison to the 3rd quartile value also suggests this.

```{r outliers, echo =FALSE}
#Define insulin variable
insulin <- diabetes_df$Insulin
#Use boxplot function to check for outliers in Insulin data by plotting a boxplot
boxplot(insulin, main = "Box Plot of Insulin")
#Use hist function to check for outliers in insulin data by plotting a histogram
hist(insulin, main = "Histogram of Insulin",  xlab = "Insulin")
```

Extreme values indicated by box plot are to be kept as they are plausible real-world values.

# Define Variables based on columns in the diabetes data frame
```{r variables, echo=FALSE}

pregnancies <- diabetes_df$Pregnancies
glucose <- diabetes_df$Glucose
blood_pressure <- diabetes_df$BloodPressure
skin_thickness <- diabetes_df$SkinThickness
bmi <- diabetes_df$BMI
diabetes_pedigree_fn <-diabetes_df$DiabetesPedigreeFunction
age <- diabetes_df$Age


for (col in colnames(diabetes_df)){
  if (col != "Outcome") {
  plot <-  ggplot(diabetes_df, aes( x = .data[[col]], y = outcome)) +
  geom_point(color = "blue", size = 2 ) +
  labs(title = paste("Plot of", col, "vs Outcome"),
       x = col,
       y = "Outcome")
  print(plot)
}}
```

The initial plots show that no. of pregnancies, glucose, blood pressure may be correlated with the outcome of diabetes, which should be validated further with logistic model applications.   

# 2. Splitting the Dataset

The dataset is split into training and test sets using an 80:20 ratio.

```{r splitting, echo=FALSE}

#Split the dataset by a ratio of 80:20
split <- sample.split(outcome, SplitRatio = 0.8)

#Training set is 80% of the dataset
train_reg <- diabetes_df[split, ]


#Test set is 20% of the dataset
test_reg <- diabetes_df[!split, ]


```

# 3. Fit Logistic Regression Model

```{r lm comment, echo=FALSE}
#Use glm function to fit a logistic regression model
```

```{r lm}

logistic_model <- glm(outcome ~ pregnancies + glucose + blood_pressure,
                    data = train_reg,
                    family = "binomial")

```
# 4.  Model Summary 
```{r lm summary , echo=FALSE}
#Use summary function to display summary statistics of the linear regression model
summary(logistic_model)

```

For no. of pregnancies and glucose levels they have p-values which are less than 0.05, suggesting they are statistically significant predictors of diabetes outcomes. Both variable coefficients of 0.123287 and 0.037080 are positive, which suggests the higher number of pregnancies and higher glucose levels increase the probability of diabetes in this dataset. 

However for blood pressure, this has a high p-value > 0.1, indicating that this is not a statistically significant indicator of the outcome. The model can be refitted to remove the variable that is not significant. 

```{r lm_v1, echo=FALSE}
#Use glm function to fit a logistic regression model
logistic_model_v1 <- glm(outcome ~ pregnancies + glucose,
                    data = train_reg,
                    family = "binomial")

#Use summary function to display summary statistics of the logistic regression model
summary(logistic_model_v1)
```

The AIC value is slightly lower at 790.95 compared to 792.06 from the first fit of the model, which suggests this a better model with the removal of the variable that is not statistically significant. 

The difference between the residual and the null deviance is equal to 208.53. The difference between the degrees of freedom between residual and null deviance is equal to 2; corresponding to the 2 additional predictors. This suggests that the logistic model reduces the deviance in the model and is a better fit compared to the null model with the addition of the pregnancy and glucose variables. 


# 5. Check Assumptions

Assumption #1: The Response Variable is Binary

Display the type of each variable:
```{r type_for_binary_test, echo =FALSE}
str(diabetes_df)
```
Assumption met: Only two possible outcomes in this data set, "Yes" and "No" as logistic regression assumes a binary response variable. 

Assumption #2: The Observations are Independent

Independence can only be verified on data collection and cannot be formally tested. 

Assumption #3: There is No Multicollinearity Among Explanatory Variables

```{r VIF, echo=FALSE}
#Calculate the VIF value of the logistic regression model

vif(logistic_model_v1)

#Create a vector value of the VIF values

vifs <- vif(logistic_model_v1)

#Display each vector value on a horizontal bar chart
barplot(vifs, main ="VIF Values", horiz = TRUE, col = "purple")

```
The VIF values for no. of pregnancies and glucose levels are 1.001006 which is less than 5, which indicates very little correlation between the two variables,  so the assumption of no multicollinearity is met for this model.

Assumption #4: There are No Extreme Outliers

```{r cd, echo=FALSE}

#find Cook's distance for each observation in the dataset
cd <- cooks.distance(logistic_model_v1)

# Plot Cook's Distance with a horizontal line at 4/n to see which observations exceed this threshold

#identify no. of row in diabetes datframe
n <- nrow(diabetes_df)
#plot cooks distance 
plot(cd, main = "Cooks Distance for Influential Obs")
#add a cutoff line at 4/n to identify influential observations
abline(h = 4/n, lty = 2, col = "steelblue") 

```

There are some observations in this data set that largely exceed the 4/n line. These datapoints are likely influencial points and have a strong influence on the regression model. As this data set represents real life data these values should remain. However, this observation does serve as a limitation of this model. 


#5: There is a Linear Relationship Between Explanatory Variables and the Logit of the Response Variable

```{r bt, echo = FALSE}

#Create a copy of the dataframe for box Tidwell
diabetes_df_bt <- copy(diabetes_df)

#Remove 0 values as box Tidwell only accepts positive values
diabetes_df_bt <- subset(diabetes_df_bt, Pregnancies > 0 & Glucose > 0)

# For each of the predictor columns create a column with contains the natural log i.e. Pregnancies * Log(Pregnancies)

diabetes_df_bt$LogPregnancies <- diabetes_df_bt$Pregnancies *log(diabetes_df_bt$Pregnancies)
diabetes_df_bt$LogGlucose <- diabetes_df_bt$Glucose * log(diabetes_df_bt$Glucose)

# Fit the logistic model again with the addition of the natural log columns

logistic_model_v2 <- glm(Outcome ~ Pregnancies + Glucose + LogPregnancies + LogGlucose,
                         data = diabetes_df_bt,
                         family = binomial)

summary(logistic_model_v2)
```
For the natural log values of no. of pregnancies and glucose levels they have p-values which are greater than 0.1; LogPregnancies (0.98127) and  LogGlucose (0.34402). Therefore, these variables are likely not statistically significant predictors of diabetes outcomes. This means that the assumption of linearity between the explanatory variables and the logit of the response variable is met. This was tested using Box-Tidwell. 

#6: The Sample Size is Sufficiently Large

# Find total number of observations
```{r len df, echo=FALSE}
# Find total number of observations
print(paste("Rows in the Dataframe :",nrow(diabetes_df)))
```
Total number of observations is 768 which is greater than 500, therefore the assumption of large sample size is met. 