---
title: "Logistic Regression Analysis Report"
author: "Grace Hardy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

# Load packages 
library(tidyverse)
library(broom)
library(ggplot2)
library(lmtest)
library(car)
library(caret)
library(MASS)
library(glmnet)
library(remotes)
library(rlang)
library(here)
library(caTools)
library(data.table)
library(InformationValue)

```

```{r fileload, include=FALSE}
# Create file path, the here function enables files to be located relative to the project root
diabetes_path <- here("Data", "diabetes.csv")

#tryCatch statement to safely try loading diabetes data
if (file.exists(diabetes_path)) {
  diabetes_df <- tryCatch(
    read.csv(diabetes_path, stringsAsFactors = FALSE),
    error = function(e) {
      stop("Error reading CSV file: ", e$message)
    }
  )
  cat("Diabetes file loaded successfully.\n")
} else {
  cat("Diabetes file not found in current working directory.\n")
}
```
# 1. Exploratory Analysis

Print the first few rows of the diabetes data:
```{r head, echo=FALSE}
# Use head function to display first 6 rows 
head(diabetes_df)
```

Print the last few rows of the diabetes data:
```{r tail, echo=FALSE}
# Use tail function to display last 6 rows 
tail(diabetes_df)
```
No obvious data or formatting errors and the column headings look representative of their values. 

Summary statistics of the diabetes data:
```{r summary, echo=FALSE}
# Use summary function to display summary statistics of data frame 
 summary(diabetes_df)
```

Display the type of each variable:
```{r type, echo =FALSE}
str(diabetes_df)
```

All variables are numerical so do not need converting to a factor except for the Outcome variable which should be treated as a factor for logistic regression. 

```{r Outcome Variable Conversion, echo=FALSE}
#Define Outcome variable
outcome <- diabetes_df$Outcome

#Convert outcome variable to factor
diabetes_df$Outcome <- factor(diabetes_df$Outcome, levels = c(0, 1), labels = c("No", "Yes"))

#Define Outcome variable
outcome <- diabetes_df$Outcome
```

For pregnancies, maximum value of 17 seems quite large but mean is quite close to the median so no indication of right-skewness and this could be a plausible real-world value.

Zero values for Glucose, BloodPressure, SkinThickness, Insulin and BMI could suggest some missing data collection. This is shown below. 

```{r missing values, echo=FALSE}

#Find all rows that have a value of 0
colSums(diabetes_df == 0)

```

The table above shows that the highest number of zero values occur in the SkinThickness and Insulin variables, further implying that there is some missing data for these values, which may impact the validity of further model applications. Pregnancy is also showing as having zero values, but this is a real-world expected value.

For Insulin, the mean is significantly larger than the median which could indicate right skewness or presence of outlier values, the large max value in comparison to the 3rd quartile value also suggests this.

```{r outliers, echo =FALSE}
#Define insulin variable
insulin <- diabetes_df$Insulin
#Use boxplot function to check for outliers in Insulin data by plotting a boxplot
boxplot(insulin, main = "Box Plot of Insulin")
#Use hist function to check for outliers in insulin data by plotting a histogram
hist(insulin, main = "Histogram of Insulin",  xlab = "Insulin")
```

Extreme values indicated by box plot are to be kept as they are plausible real-world values.

# Define Variables based on columns in the diabetes data frame
```{r variables, echo=FALSE}

pregnancies <- diabetes_df$Pregnancies
glucose <- diabetes_df$Glucose
blood_pressure <- diabetes_df$BloodPressure
skin_thickness <- diabetes_df$SkinThickness
bmi <- diabetes_df$BMI
diabetes_pedigree_fn <-diabetes_df$DiabetesPedigreeFunction
age <- diabetes_df$Age


for (col in colnames(diabetes_df)){
  if (col != "Outcome") {
  plot <-  ggplot(diabetes_df, aes( x = .data[[col]], y = outcome)) +
  geom_point(color = "blue", size = 2 ) +
  labs(title = paste("Plot of", col, "vs Outcome"),
       x = col,
       y = "Outcome")
  print(plot)
}}
```

The initial plots show that no. of pregnancies, glucose, blood pressure may be correlated with the outcome of diabetes, which should be validated further with logistic model applications.   

# 2. Splitting the Dataset

The dataset is split into training and test sets using an 80:20 ratio.

```{r splitting, echo=FALSE}

#Split the dataset by a ratio of 80:20
split <- sample.split(outcome, SplitRatio = 0.8)

#Training set is 80% of the dataset
train_reg <- diabetes_df[split, ]


#Test set is 20% of the dataset
test_reg <- diabetes_df[!split, ]


```

# 3. Fit Logistic Regression Model

```{r lm comment, echo=FALSE}
#Use glm function to fit a logistic regression model
```

```{r lm}

logistic_model <- glm(Outcome ~ Pregnancies + Glucose + BloodPressure,
                    data = train_reg,
                    family = "binomial")

```
# 4.  Model Summary 
```{r lm summary , echo=FALSE}
#Use summary function to display summary statistics of the linear regression model
summary(logistic_model)

```

For no. of pregnancies and glucose levels they have p-values which are less than 0.05, suggesting they are statistically significant predictors of diabetes outcomes. Both variable coefficients of 0.134309 and 0.039849 are positive, which suggests the higher number of pregnancies and higher glucose levels increase the probability of diabetes in this dataset. 

However for blood pressure, this has a high p-value > 0.05, indicating that this is not a statistically significant indicator of the outcome. The model can be refitted to remove the variable that is not significant. 

```{r lm_v1, echo=FALSE}
#Use glm function to fit a logistic regression model
logistic_model_v1 <- glm(Outcome ~ Pregnancies + Glucose,
                    data = train_reg,
                    family = "binomial")

#Use summary function to display summary statistics of the logistic regression model
summary(logistic_model_v1)
```

The AIC value is actually higher at 625.53 for the second fit of the model compared to 624.65 from the first fit of the model. Lower AIC scores indicate a better fit of the model which suggests the first model with the inclusion of the blood pressure model is a better fit. 

The difference between the residual and the null deviance is equal to 177.29. The difference between the degrees of freedom between residual and null deviance is equal to 3; corresponding to the 3 additional predictors. This suggests that the logistic model reduces the deviance in the model and is a better fit compared to the null model with the addition of the pregnancy, glucose, blood pressure variables. 


# 5. Check Assumptions

Assumption #1: The Response Variable is Binary

Display the type of each variable:
```{r type_for_binary_test, echo =FALSE}
str(diabetes_df)
```
Assumption met: Only two possible outcomes in this data set, "Yes" and "No" as logistic regression assumes a binary response variable. 

Assumption #2: The Observations are Independent

Independence can only be verified on data collection and cannot be formally tested. 

Assumption #3: There is No Multicollinearity Among Explanatory Variables

```{r VIF, echo=FALSE}
#Calculate the VIF value of the logistic regression model

vif(logistic_model)

#Create a vector value of the VIF values

vifs <- vif(logistic_model)

#Display each vector value on a horizontal bar chart
barplot(vifs, main ="VIF Values", horiz = TRUE, col = "purple")

```

The VIF values for no. of pregnancies (1.025205), glucose (1.026410) and blood pressure (1.051884) levels are all around 1 which is far less than the threshold, which indicates very little correlation between the 3variables, so the assumption of no multicollinearity is met for this model.

Assumption #4: There are No Extreme Outliers

```{r cd, echo=FALSE}

#find Cook's distance for each observation in the dataset
cd <- cooks.distance(logistic_model)

# Plot Cook's Distance with a horizontal line at 4/n to see which observations exceed this threshold

#identify no. of row in diabetes datframe
n <- nrow(train_reg)
#plot cooks distance 
plot(cd, main = "Cooks Distance for Influential Obs")
#add a cutoff line at 4/n to identify influential observations
abline(h = 4/n, lty = 2, col = "steelblue") 

```

There are some observations in this data set that largely exceed the 4/n line. These datapoints are likely influencial points and have a strong influence on the regression model. As this data set represents real life data these values should remain. However, this observation does serve as a limitation of this model. 


#5: There is a Linear Relationship Between Explanatory Variables and the Logit of the Response Variable

```{r bt, echo = FALSE}

#Create a copy of the dataframe for box Tidwell
train_reg_bt <- copy(train_reg)

#Remove 0 values as box Tidwell only accepts positive values
train_reg_bt <- subset(train_reg_bt, Pregnancies > 0 & Glucose > 0 & BloodPressure > 0)

# For each of the predictor columns create a column with contains the natural log i.e. Pregnancies * Log(Pregnancies)

for (col in c("Pregnancies", "Glucose", "BloodPressure")){
  new_col_name = paste("Log",col, sep="")
  train_reg_bt[[new_col_name]] <- train_reg_bt[[col]]*log(train_reg_bt[[col]])
}

# Fit the logistic model again with the addition of the natural log columns

logistic_model_v2 <- glm(Outcome ~ Pregnancies + Glucose + BloodPressure + LogPregnancies +                          LogGlucose + LogBloodPressure, 
                         data = train_reg_bt,
                         family = binomial)

summary(logistic_model_v2)
```
For the natural log values of no. of pregnancies and glucose levels they have p-values which are greater than 0.1; LogPregnancies (0.3414), LogGlucose (0.2415) and LogBloodPressure (0.9498). Therefore, these variables are likely not statistically significant predictors of diabetes outcomes. This means that the assumption of linearity between the explanatory variables and the logit of the response variable is met. This was tested using Box-Tidwell. 

#6: The Sample Size is Sufficiently Large

# Find total number of observations
```{r len df, echo=FALSE}
# Find total number of observations
print(paste("Rows in the Dataframe :",nrow(diabetes_df)))
```
Total number of observations is 768 which is greater than 500, therefore the assumption of large sample size is met. 


# 6. Make Predictions 


```{r predictions, echo= FALSE}
#define an individual with a value for no. of pregnancie, glucose and blood pressure levels 
new <- data.frame(Pregnancies = 1, Glucose = 175, BloodPressure = 120)

#predict probability of diabetes outcome 
predicted_new <- predict(logistic_model, new, type="response")
print (predicted_new)


#calculate probability of diabetes outcome for each individual in test dataset
predicted <- predict(logistic_model, test_reg, type="response")


```

The prediction is based on the first fit of the model: logistic model due to this model having a lower AIC score. The probability of an individual who has had 1 pregnancy, has glucose levels of 175 and blood pressure levels of 120 has a probability of having diabetes of 0.6819376.

# 7. Model Diagnostics 

```{r diagnostics, echo=FALSE}
#Convert the Outcome Values from "Yes" / "No" back to binary values
test_reg$Outcome <- ifelse(test_reg$Outcome=="Yes", 1, 0)

#find optimal cutoff probability to use to maximise accuracy
optimal <- optimalCutoff(test_reg$Outcome, predicted)[1]
optimal

```

The optimal value cutoff to use is 0.556254. This means, any individual with a probability of having diabetes of 0.5562548 or higher will be predicted to have diabetes. Those with a probability less than 0.5562548 will be predicted to not.

```{r confusion matrix, echo = FALSE}
#Create a confusion matrix 
confusionMatrix(test_reg$Outcome, predicted)

#Calculate the sensitivity (true positive rate)
sensitivity(test_reg$Outcome, predicted)

#Calculate the specificity (true negative rate)
specificity(test_reg$Outcome, predicted)

#Calculate total misclassification error rate (% of total incorrect classifications)
misClassError(test_reg$Outcome, predicted, threshold=optimal)

```

True postive rate of 42.6% indicates that the model correctly predicts a positive diabetes outcome in 42.6% of cases. For the true negative rate, the model accurately predicts a negative diabetes outcome in 88% of cases. Misclassification error is 24.03 %, which is quite high, this indicates that the model may have some limitations in accurate prediction of whether an individual will have diabetes or not, this could perhaps be influenced by the low sensitivity rate.

```{r ROC, echo=FALSE}

#plot the ROC curve
plotROC(test_reg$Outcome, predicted)

```

The AUC is 0.8363, which is reasonably high, indicates the model does a fairly good job of predicting the diabetes outcome. 


# 7. Conclusion 

The logistic regression model with predictors no. of pregnancies, glucose and blood pressure levels is fairly good at predicting the diabetes outcome as shown by its AUC value. However, this is limited due to the violation of the assumption that there are no extreme outliers as shown by the cooks distance test as well as the fairly high misclassification error and low sensitivity.

